---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# MarineTides

<!-- badges: start -->
[![License: GPL3](https://img.shields.io/badge/License-GPL3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
<!-- badges: end -->

## Introduction

The `MarineTides` package can be used to generate tide predictions for 3,333 tide stations worldwide. The majority of these stations are located in the United States. Remaining stations are located in areas adjacent to American territories in the Pacific Ocean and Carribbean. This includes places such as the Marshall Islands, Fiji, Cuba, and the Virgin Islands. Both harmonic and subordinate tide stations are included. At current count, there are 1,143 stations with harmonic constituents, and 2,190 subordinate stations. 

The package name is intended to distinguish Marine Tides from [`Earth Tides`](https://en.wikipedia.org/wiki/Earth_tide), or tides in large freshwater bodies such as the [`Great Lakes`](https://oceanservice.noaa.gov/facts/gltides.html). The `MarineTides` package can only be used to predict tides for stations in coastal locations, the open ocean, or marine influenced waters such as estuaries. 

All station data, including harmonic constituents, were downloaded from the [NOAA CO-OPS API for Data Retrieval](https://api.tidesandcurrents.noaa.gov/api/prod/#products). Lunar node year corrections to tidal constituents for years 1700 to 2100 were extracted from the output file of David Flater's [congen program](https://flaterco.com/xtide/files.html#harmonicsfiles). The node year data captures the effects of the 18.6 year lunar orbital cycle on amplitudes and phase of lunar tidal constituents. In order to match NOAA CO-OPS harmonic constituent naming conventions, some constituent names needed to be updated. Issues related to constituent naming conventions, and derivation of lunar nodal corrections are documented by David Flater in the [`congen`](https://flaterco.com/files/xtide/congen_input.txt) program input file.

For tide stations where [NOAA CO-OPS](https://api.tidesandcurrents.noaa.gov/api/prod/#products) provide harmonics data, predictions can be made at intervals from one minute to one hour. This is modeled on the options available at [NOAA Tides & Currents](https://tidesandcurrents.noaa.gov/) Alternately, values can also be generated for just daily high and low tides. 

For subordinate stations, predictions are restricted to daily highs and lows. These subordinate station predictions are based on tide levels calculated for an adjacent harmonic reference station. The reference station predictions are generated at one-minute intervals. Then local peaks and valleys in the harmonic tidal curves are extracted to obtain the daily highs and lows. Finally, offset values provided by NOAA CO-OPS are used to adjust the harmonic derived values and generate subordinate station estimates. 

## Motivation

The `MarineTides` package is motivated by need for large volumes of tide predictions, often at one-minute increments. Primary uses are for intertidal shellfisheries management, and for climate-related marine research. The [`rtide`](https://github.com/poissonconsulting/rtide) package has ably filled this need in the past, but speed can become an issue when longer time-series at higher resolutions are needed. NOAA sensibly limits the amount of data that can be downloaded in one batch, and `rtide` can take several minutes to churn out a full years worth of predictions at one-minute intervals. Another limitation of `rtide` is that it does not allow for tide predictions at subordinate stations. 

In addition, upcoming revisions of the [National Tidal Datum Epoch](https://tidesandcurrents.noaa.gov/datum-updates/ntde/) and the [North American Vertical Datum (NAVD88)](https://oceanservice.noaa.gov/geodesy/three-datums.html), suggests the need for a tide prediction package that can be easily updated as new datums and harmonic data become available. Minor changes to station data commonly occur on a quarterly schedule, and updates to datums will likely occur more frequently as climate warms and sea levels rise. The aim is to update `MarineTides` on a semi-regular schedule to reflect changes to station data pushed by [`NOAA CO-OPS`](https://tidesandcurrents.noaa.gov/products.html). 

`MarineTides` has intentionally been designed with minimal dependencies and makes use of functions in the [`data.table`](https://rdatatable.gitlab.io/data.table/) package to speed up calculations. [NOAA CO-OPS](https://tidesandcurrents.noaa.gov/about_harmonic_constituents.html) has an excellent write-up of the formula and methods used to predict tides. The core functions in `MarineTides` were derived in part from Joe Thorley's `rtide` package. In turn, `rtide` was modeled on David Flater's [`XTide`](https://flaterco.com/xtide/).  

## Installation

`MarineTides` is still in an early stage of development and some functions may change as needs arise. The core API will hopefully remain stable, but there are no guarantees. Expect surprises! Do not use for navigation!

To install the latest version from [GitHub](https://github.com/arestrom/MarineTides)
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("arestrom/MarineTides")
```

## Contributions

Please report any [issues](https://github.com/arestrom/MarineTides/issues).

[Pull requests](https://github.com/arestrom/MarineTides/pulls) are always welcome.
