---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# MarineTides

<!-- badges: start -->
[![License: GPL3](https://img.shields.io/badge/License-GPL3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0.en.html)
<!-- badges: end -->

## Introduction

`MarineTides` is a package of functions that can be used to generate tide predictions for over three thousand locations worldwide. The majority of these stations are located in the United States, but others are located in areas adjacent to American territories in the Pacific Ocean and Carribbean. This includes places such as the Marshall Islands, Fiji, Cuba, and the Virgin Islands. 

For tide stations where [NOAA CO-OPS](https://api.tidesandcurrents.noaa.gov/api/prod/#products) have calculated harmonics data, predictions can be made at intervals from one minute to one hour. Alternately, values for just the daily high and low tides can be requested. 

For subordinate stations, predictions are restricted to the daily high and low tide values. These subordinate predictions are based on tide levels calculated for an adjacent harmonic reference station. The reference station predictions are generated at one-minute intervals. Then local peaks and valleys in the harmonic tidal curves are extracted to obtain the daily highs and lows. Finally, offset values published by NOAA CO-OPS are used to adjust the harmonic values and generate subordinate station estimates. 

## Example

```{r}
library(MarineTides)

# Whitney Point is a subordinate station located on Hood Canal, Washington State. USA.
data <- MarineTides::tide_level(
  station = "Whitney Point",
  start_date = "2004-02-04",
  end_date = "2004-02-05",
  data_interval = "high-low"
)

print(data)
```

## Installation

`MarineTides` is still in an early stage of development. It has not been fully tested yet. You should expect surprises!

To install the latest version from [GitHub](https://github.com/arestrom/MarineTides)
```{r, eval=FALSE}
# install.packages("remotes")
remotes::install_github("arestrom/MarineTides")
```

## Contributions

Please report any [issues](https://github.com/arestrom/MarineTides/issues).

[Pull requests](https://github.com/arestrom/MarineTides/pulls) are always welcome.

## Inspiration

Almost since the first release, Joe Thorley's excellent [`rtide` package](https://github.com/poissonconsulting/rtide) has been an indispensible tool to generate tides for research and analysis. But speed can become an issue when longer time-series and higher resolutions are needed. NOAA sensibly limits the amount of data that can be downloaded in one batch, and `rtide` can take several minutes to churn out a full years worth of predictions. Another limitation of `rtide` is that it does not allow tide predictions for subordinate stations. 

In addition, upcoming revisions of the [National Tidal Datum Epoch](https://tidesandcurrents.noaa.gov/datum-updates/ntde/) and the [North American Vertical Datum (NAVD88)](https://oceanservice.noaa.gov/geodesy/three-datums.html), suggested the need for a new package that I could quickly adapt as new datums and harmonic data became available. Updates to datums will likely occur more frequently as climate warms and sea levels rise. 

`MarineTides` has intentionally been designed with minimal dependencies and makes use of functions in the `data.table` package to speed up calculations. [NOAA CO-OPS](https://tidesandcurrents.noaa.gov/about_harmonic_constituents.html) has an excellent write-up of the formula and methods used to predict tides. The core functions in `MarineTides` were derived in large part from the `rtide` package. In turn, `rtide` was modeled on David Flater's [`XTide`](https://flaterco.com/xtide/).  

